% contextualization
In order to prove the correctness of our \btrees\ library, we first need a formal model, in Coq, for each type and function.
This model will be used to specify each C function with VST.
Then, the correctness can be proved by showing that the functions of the formal model comply with the formal specification, thus leveraging the proof to a Coq one without C semantics.

% why not Brian's model ?
DeepDB already contained a formal model that complied with the abstract relation axiomatization~\cite{brian}.
However, this \btree\ model does not exactly mimic the behavior of the C code.
For instance, entries of the C implementation contain a pointer to the child with greater keys.
Then, the node contains a pointer to the first child.
However, in this formal model, each entry contains a node which keys are lesser, and the node holds the pointer to the last child.
Moreover, the \texttt{First} and \texttt{Last} booleans used to speed up the functions are not present in each node, and not used in the functions.
As a result, we decided to define another formal model, as close as possible to the C code.
This new formal model will be used to specify the C functions with VST.


% Types
\begin{figure}
\begin{lstlisting}[language=Coq]
(* Btree Types *)
Inductive entry (X:Type): Type :=
     | keyval: key -> V -> X -> entry X
     | keychild: key -> node X -> entry X
with node (X:Type): Type :=
     | btnode: option (node X) -> listentry X -> bool -> bool -> bool -> X -> node X
with listentry (X:Type): Type :=
     | nil: listentry X
     | cons: entry X -> listentry X -> listentry X.

Definition cursor (X:Type): Type := list (node X * index). (* ancestors and index *)
Definition relation (X:Type): Type := node X * X.  (* root and address *)
\end{lstlisting}
\label{coqtypes}
\caption{Coq Formal Model Types}
\end{figure}

% Types explanation
\paragraph{Types}\textbf{Fig.}~\ref{coqtypes} presents the Coq Types for \btrees, cursors and relations. \todo{update for dependent relation type}
These types are parametrized by a type \texttt{X}, that can be either \texttt{val} or \texttt{unit}.
Using \texttt{val} allows to add the C address of each node, record or relation directly into the formal type.\todo{why?}
We can see that entries have a key, and either a record (typed \texttt{V}) or a child (typed \texttt{node X}).
Nodes have three booleans, representing the \texttt{isLeaf}, \texttt{First} and \texttt{Last} of the C code.
The $ptr_0$ of a node is represented with an option, as Leaf Nodes don't have any.

A cursor is implemented in Coq as a list of nodes and indexes.
This corresponds to the arrays found in the C code. The list starts at the root and its head is the current node and index.
Its length indicates the cursor's depth.
Finally, a relation is simply a root (node) and the address at which the representation is in the memory.% The depth and number of records can be easily computed.

\paragraph{Functions} Then, each C function must have an equivalent in the formal Coq model.
For instance, \textbf{Fig.}~\ref{movetofirst} presents the coq \texttt{moveToFirst} function.
It takes as input the next node to go down to. If this node is a Leaf Node, then it returns the cursor with the new node, and the index 0 (pointing to the first record).
Otherwise, it goes down $ptr_0$, and adds to the cursor the next node and the index \texttt{im} (representing -1).

\begin{figure}
\begin{lstlisting}[language=Coq]
(* takes a PARTIAL cursor, n next node (pointed to by the cursor)
   and goes down to the first key *)
Fixpoint moveToFirst {X:Type} (n:node X) (c:cursor X) (level:nat): cursor X :=
  match n with
    btnode ptr0 le isLeaf First Last x =>
    match isLeaf with
    | true => (n,ip 0)::c
    | false => match ptr0 with
               | None => c      (* not possible, isLeaf is false *)
               | Some n' => moveToFirst n' ((n,im)::c) (level+1)
               end
    end
  end.
\end{lstlisting}
\caption{MoveToFirst in the formal model}
\label{movetofirst}
\end{figure}  
