% deepdb
This verification work is part of the DeepDB project\todo{ref}.
Its goal is to provide a verified library for a high-performance concurrent Main-Memory database.
Inspired by MassTree, this database system also uses a combination of \btrees\ and Tries structures.

% what has been done
As part of the deepDB project, a formal Coq specification of abstract relations with cursors has been defined\todo{ref thesis Brian}.
A C implementation of a sequential version of the database has also been done\todo{ref thesis Tosin}.
This implementation contains a first version of a library for \btrees\ wih cursors, a library for border nodes~\cite{masstree} and a library for Tries.
This work is the first verification step: it includes modifying the \btrees\ library and proving it correct with regards to the specification.

% verif of trees
Previous work has dealt with the verification of tree-like structures.
Verified Functional Algorithms~\cite{vfa} includes a Coq formal model for Binary-Search Trees and its verification.
Red-Black Trees have also been verified in~\cite{redblack}.
VST's examples~\cite{vst} include the VST verification of a C implementation of Binary-Search Trees.

% concurrent
Even though the current version is sequential, we believe that implementing concurrency would be a feasible next step.
Verification of concurrent C program has been considered in the past\todo{refs?}.
Recent work in VST allows to prove correctness of concurrent C programs\todo{ref?}.

\todo{...}

